Epistemic Semantics in Guarded String Models

### 2020-09-19 tex/sdraft.tex
Submission dradt sdraft.

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a
natural language.  A proposition is a set or class of {\em possible
  worlds}.  These are often assumed to be things with the same nature
and complexity as the world we occupy \citep{lewis1986plurality}.  But
alternatively, one can work with small idealized models, in order to
illustrate and test ideas.  To build such models, spaces of worlds and
individuals are stipulated as small finite sets, and semantic values
of lexical items are constructed as functions or relations from these
sets.  Such toy or idealized models are useful in research and in
teaching, in that it is possible to represent propositions finitely
and explicitly, and to calculate with them. The point of this paper is
to scale up toy or idealized models to countable sets of worlds, and
to constructive and computable modeling of epistemic alternatives for
agents.  We describe a certain systematic way of defining such models,
and illustrate how to apply them in natural language semantics.  The
focus is on epistemic semantics and clausal embedding. The fundamental
move is to identify possible worlds with strings of primitive events,
so that propositions are sets of strings. An advantage in this is that
it allows for a mathematical description of an algebra of
propositions, coupled with a computational representation using either
lazy lists of strings, or finite state machines that describe sets of
strings.

@@

We begin with examples of event-sequence models.  {\em The
  Elevator}. An elevator moves up and down in a four-story building,
with floors numbered in the European fashion as 0,1,2,3.  There are
primitive events $u$ (the elevator going up one floor), and $d$ (the
elevator going down on floor).  In worlds $v_1$ and $v_2$, the events
shown in \ex{elevator1} transpire. The truth values for English
sentences shown in \ex{elevator2} are observed.

\begin{example}  \label{elevator1}
\hspace{-4mm} $v_1$
\begin{tabular}[t]{ll}
  $u$ & it goes up from 0 to 1 \\
  $u$ & it goes up from 1 to 2 \\
  $d$ & it goes down from 2 to 1 \\
  $u$ & it goes up from 1 to 2 \\
\end{tabular}

\vspace{2mm}
\hspace{-4mm} $v_2$
\begin{tabular}[t]{ll}
  $u$ & it goes up from 0 to 1 \\
  $u$ & it goes up from 1 to 2 \\
  $u$ & it goes up from 2 to 3 \\
\end{tabular}
\end{example}

\begin{example}  \label{elevator2}
\hspace{-4mm} \begin{tabular}[t]{ccl}
  $v_1$ & $v_2$ & Sentence \\
   false & true & It's on floor 3. \\
   true & true & It has gone up. \\
   true & false & It has gone down. \\
   true & false & It could go up. \\
\end{tabular}
\end{example}

{\em The Concealed Coin}.  Amy and Bob are seated at a table.  There is a coin on the
table under a cup, heads up.  The coin could be heads or tails, and
neither agent knows which it is.  This initial situation is possible world $w_1$.
Two additional worlds $w_2$ and $w_2$ are
defined by sequencing events after the initial state, with events interpreted as in
\ex{coin1}. The truth values for English
sentences shown in \ex{coin3} are observed.

@@

The events in the examples come with pre-conditions. The elevator can
not go up if it is already on floor 3, so $u$ has the pre-condition of
the elevator being of floor 0, 1, or 2.  Similarly $d$ has the
precondition that the elevator is on floor 1, 2 or 3.  Amy can peek at
heads only if the coin is heads up, so $a_1$ has the precondition of
the coin being heads up.  Let $h$ be the Boolean proposition that the
coin is heads up.  In the other example, let $q$ be the proposition
that the elevator is on a high floor (2 or 3), and $p$ be the
proposition that it is on an odd floor (1 or 3). Then preconditions
can be described by Boolean formulas, with $h$ being the precondition
of $a_1$, and $\overline{pq}$ being the precondition of
$u$. Juxtaposition is used for Boolean conjunction, and the overbar
for Boolean negation. Events come as well with a relation between
prior and following state, for instance with $u$ incrementing the
floor.  This is expressed using an operator ``$:$'' (read ``and
next'') that pairs Boolean formulas. The first line in \ex{event1}
describes $a_1$ (Amy looking at heads) as happening only in an $h$
state, and as not changing the state.  Symmetrically, $a_0$ (Amy
looking at tails) can happen only in a not-$h$ state, and does not
change the state.  The third line says that $u$ increments the floor,
and can happen only on floors 0, 1, and 2.  The fourth line describes
$d$ in similar terms.  Plus is disjunction.

@@

\begin{example}   \label{event1}
\hspace{-5mm}\( \begin{array}[t]{cl}
a_1 & h:h \\
a_0 & \ev{\bar{h}}{\bar{h}} \\
% u & (\bar{q}\bar{p}) \! : \!(\bar{q}p) + \\
u & \ev{(\bar{q}\bar{p})}{(\bar{q}p)} + \ev{(\bar{q}p)}{(q\bar{p})} + \ev{(q\bar{p})}{(qp)} \\
d & \evr{(\bar{q}\bar{p})}{(\bar{q}p)} + \evr{(\bar{q}p)}{(q\bar{p})} + \evr{(q\bar{p})}{(qp)} \\ 
\end{array} \)
\end{example}

@@

In the discussion at the end of the last section, a sequence such as
$\bar{q}{p}$ can be viewed as a valuation of primitive test
propositions, which is used to describe world state. The primitives
are listed in fixed order, and left unmarked (indicating true) or
marked with the overbar (indicating false).  Suppose that in the coin
example, we have an additional primitive stative proposition (or
atomic test) $t$, interpreted as tails.  Since a coin is heads or
tails but not both, we want to allow the valuations $h\bar{t}$ and
$\bar{h}t$, and disallow $ht$ and $\bar{h}\bar{t}$.  This is enforced
by a {\em state formula}, which is a Boolean formula, in this case the
one given on the second line of \ex{effect}.  Where {\sf{B}} is a set
of atomic tests and $\phi$ is a state constraint over {\sf{B}},
${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\phi}$ is the set of valuations of
{\sf{B}} that make formula $\phi$ true. Valuations are called atoms,
because they correspond to the atoms of a Boolean algebra of tests
\citep{kozen2001automata}.

@@

\begin{example}   \label{gs1}
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}[t]{clc}
{\textit{World}} & {\textit{Guarded string}} & \textit{Length}\\ \hline
$v_1$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \,u\, \hat 2$ & $4$ \\
$v_2$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,u\, \hat 3$ & $3$ \\   
$w_1$ & $H$ & $0$\\
$w_2$ & $H \,a_1\, H$ & $1$ \\  
$w_3$ & $H\,a_1\, H \,b_1\, H$ & $2$\\
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

@@

\begin{example} \label{fusion1}
\hspace{-2mm}\(\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \; \diamond \; \hat 1 \,u\, \hat 2 = \hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \,u\, \hat 2 \)

\hspace{-2mm}\( \hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \; \diamond \; \hat 2 \,u\, \hat 3 = \textit{undefined} \)

@@

\hspace{2mm}\(\sem{ \Diamond_a\,p } \hspace{4mm}{\triangleq} \hspace{4mm}\{x \mid \exists y . x{\hat{R}_a}y \wedge y \in \sem{ p }^{\epikspec}\} \)
  \caption{Interpretation of Epik terms as sets of guarded strings (Version M)}
  \label{fig:interp2}
\end{figure}

\begin{figure}
  \[\begin{array}{l >{\triangleq}c l}
  \sem{ 0 }^{\varphi,{\mathcal{E}}} && \{\} \\  
  \sem{ 1 }^{\varphi,{\mathcal{E}}} && \mathcal A^\varphi_{\mathsf B}\\
  \sem{ e }^{\varphi,{\mathcal{E}}}&& \hat{\mathcal E}^\varphi(e) \\
  \sem{ \psi }^{\varphi,{\mathcal{E}}} &&  \mathcal A^{\varphi~\psi}\\
  \sem{{p+q}}^{\varphi,{\mathcal{E}}} && \sem{{p}}^{\varphi,{\mathcal{E}}} \cup \sem{ q }^{\varphi,{\mathcal{E}}}\\
  \sem{{p;q}}^{\varphi,{\mathcal{E}}} && \sem{ p }^{\varphi,{\mathcal{E}}} \diamond \sem{{q}}^{\varphi,{\mathcal{E}}}\\
  \sem{ p^*}^{\varphi,{\mathcal{E}}} && \displaystyle\bigcup_n {\underbrace{\sem{ p }^{\varphi,{\mathcal{E}}}\diamond \ldots \diamond \sem{ p }^{\varphi,{\mathcal{E}}}}_{n}} \\
  \sem{ \neg p }^{\varphi,{\mathcal{E}}} && \sem{ \_^*}^{\varphi,{\mathcal{E}}} \setminus \sem{ p }^{\varphi,{\mathcal{E}}} \\
  \sem{ \Diamond_a~p } && \{x \mid x \mathbin{\hat{R_a}} y, y \in \sem{ p }^{\varphi,\mathcal E}\}

  \end{array}
  \]
  \caption{Interpretation of Epik terms as sets of guarded strings}
  \label{fig:interp}
\end{figure}

@@

Other operations in the guarded string algebra as defined in FST are simpler.
Union is union in the string algebra. The complement of a proposition
is complement relative to the set of worlds, as defined by
set difference in the string algebra.

This scheme provides for an interpretation of the language of Epik
terms that is defined in Figure \ref{termfigure}.  An Epik specifications such as
Figure 1 is translated to a straight-line program of the finite state
calculus that defines constants and functions, including the ones from
Figures 2 and 3.  In Xfst or Foma, which are interpreters for the
finite state calculus, the program is read, and then propositional
terms can be mapped to finite state machines that represent sets of
guarded strings, interpreted as propositions.

@@

Page breakdown

\begin{tabular}{lll}
  1,2 & 3.75 \\
  3   & 1.25 & FST translation\\
  4   & 1.25 & Haskell Epik \\
  5   & 0.75 & English CG fragment \\
  6   & 1.0  & Examples and discussion\\
\end{tabular}

%% \begin{example}   \label{Rhat}
%% \hspace{-2mm} \(
%% \hat{R} \triangleq
%% \{\tuple{u,v} \mid \exists c \exists d. c\mathrel R d \wedge u \in \hat{c}  \wedge v \in \hat{d} \, \} \)
%% \end{example}

{\em This stuff was at the end of Section 2, it became redundant. Also it refers to the interpretation function, which is introduced in Section 3.}

\begin{example}   \label{Rhathat}
\hspace{-2mm}\(
\hat{R} \triangleq
\hspace{-2mm}\begin{array}[t]{l}
\{ \tuple{x,y} \mid \exists u_1 ... \exists u_n \exists v_1 ... \exists v_n. \\
\hspace{6mm} u_1\mathrel{R}v_1 \wedge ... \wedge u_n\mathrel{R}v_n \wedge\\
\hspace{6mm} x \in \sem{ u_1 ; ... ; u_n }^{\varphi,{\mathcal{E}}}\hspace{1mm} \wedge \\
\hspace{6mm} y \in \sem{ v_1 ; ... ; v_n }^{\varphi,{\mathcal{E}}} \hspace{2mm} \}
\end{array}
\)
\end{example}

This defines an epistemic alternative to world $x$ to be a world of
the same length, where each component event in the alternative is an
event-alternative to the event in corresponding position in the base
world. Fusion product enforces preconditions and a correspondence
between pre-states and post-states of events on both sides of the
epistemic alternative relation.  This provides for finitely
specifiable construction of epistemic models that reflect intuitions
about information exchange and epistemic consequences of perceptual
events.  See Section 6 for linguistic examples.  Since an epistemic
alternative has the same lenght as its base world, it follows from the
construction that agents know how many events have transpired in their
base worlds.

Fortunately, Epik provides a syntax to specify the input-output
behavior via the \texttt{event} declarations. From these we construct
a function $\mathcal E$ from agents to relations on atoms that specify
the effects each action can have. Notably for every declaration
$\texttt{event}~e~\zeta$, we have $\mathcal E^\varphi(e) =
\sem{\zeta}^\varphi$. We can lift $\mathcal E^\varphi$
to $\hat{\mathcal E}$ which decorates the events with their permitted
atoms, that is $\hat{\mathcal E}^\varphi(e) = \{\alpha~e~\beta \mid
\alpha, \beta \in \mathcal E^\varphi (e) \}$. See the examples in
\ex{events1}.
% u (\bar{q}\bar{p}):(\bar{q}p) + (\bar{q}p):(q\bar{p}) + (q\bar{p}):(qp) 173
% d (\bar{q}p):(\bar{q}\bar{p}) + (q\bar{p}):(\bar{q}p) + (qp):(q\bar{p})

>>>>

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a
natural language.  A proposition is a set or class of {\em possible
  worlds}.  These are often assumed to be things with the same nature
and complexity as the world we occupy \citep{lewis1986plurality}.  But
alternatively, one can work with small idealized models, in order to
illustrate and test ideas.  The point of this paper is to scale up toy
or idealized models to countable sets of worlds, and to constructive
and computable modeling of epistemic alternatives for agents.  We
describe a certain systematic way of defining such models, and
illustrate how to apply them in natural language semantics.  The focus
is on epistemic semantics and clausal embedding. The fundamental move
is to identify possible worlds with strings of primitive events, so
that propositions are sets of strings. An advantage in this is that it
allows for a mathematical description of an algebra of propositions,
coupled with a computational representation using either lazy lists of
strings, or finite state machines that describe sets of strings.

@@

A running example of an event-sequence model is {\em The Concealed
  Coin}.  Amy and Bob are seated at a table.  There is a coin on the
table under a cup, heads up.  The coin could be heads or tails, and
neither agent knows which it is.  This initial situation is possible
world $w_1$.  Two additional worlds $w_2$ and $w_2$ are defined by
sequencing events after the initial state, with events interpreted as
in \ex{coin1}. The truth values for English sentences shown in
\ex{coin3} are observed.

@@

The events come with pre-conditions.
Amy can peek at heads only if the coin is heads up, so $a_1$ has the precondition of
the coin being heads up.  Let $h$ be the Boolean proposition that the
coin is heads up.
Then preconditions
can be described by Boolean formulas, with $h$ being the precondition
of $a_1$
$u$. Events come as well with a relation between
prior and following state, for instance with $u$ incrementing the
floor.  This is expressed using an operator ``$:$'' (read ``and
next'') that pairs Boolean formulas. The first line in \ex{event1}
describes $a_1$ (Amy looking at heads) as happening only in an $h$
state, and as not changing the state.  Symmetrically, $a_0$ (Amy
looking at tails) can happen only in a not-$h$ state, and does not
change the state.

@@

\begin{example}   \label{event1}
\hspace{-5mm}\( \begin{array}[t]{cl}
a_1 & h:h \\
a_0 & \ev{\bar{h}}{\bar{h}} \\
% u & (\bar{q}\bar{p}) \! : \!(\bar{q}p) + \\
% u & \ev{(\bar{q}\bar{p})}{(\bar{q}p)} + \ev{(\bar{q}p)}{(q\bar{p})} + \ev{(q\bar{p})}{(qp)} \\
% d & \evr{(\bar{q}\bar{p})}{(\bar{q}p)} + \evr{(\bar{q}p)}{(q\bar{p})} + \evr{(q\bar{p})}{(qp)} \\ 
\end{array} \)
\end{example}

@@

Suppose that in the coin example, we have an additional primitive
stative proposition (or atomic test) $t$, interpreted as tails.  A
sequence such as $\bar{h}{t}$ can be viewed as a valuation of
primitive propositions, which is used to describe world state. The
primitives are listed in fixed order, and left unmarked (indicating
true) or marked with the overbar (indicating false).  Since a coin is
heads or tails but not both, we want to allow the valuations
$h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and $\bar{h}\bar{t}$.
This is enforced by a {\em state formula}, which is a Boolean formula,
in this case the one given on the second line of \ex{effect}.  Where
{\sf{B}} is a set of atomic tests and $\phi$ is a state constraint
over {\sf{B}}, ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\phi}$ is the set of
valuations of {\sf{B}} that make formula $\phi$ true. Valuations are
called atoms, because they correspond to the atoms of a Boolean
algebra of tests \citep{kozen2001automata}.

@@

\begin{example}   \label{gs1}
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}[t]{clc}
{\textit{World}} & {\textit{Guarded string}} & \textit{Length}\\ \hline
%$v_1$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \,u\, \hat 2$ & $4$ \\
%$v_2$ & $\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,u\, \hat 3$ & $3$ \\   
$w_1$ & $H$ & $0$\\
$w_2$ & $H \,a_1\, H$ & $1$ \\  
$w_3$ & $H\,a_1\, H \,b_1\, H$ & $2$\\
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

@@

\begin{example} \label{fusion1}
%\hspace{-2mm}\(\hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \; \diamond \; \hat 1 \,u\, \hat 2 = \hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \,u\, \hat 2 \)

%\hspace{-2mm}\( \hat 0 \,u\, \hat 1 \,u\, \hat 2 \,d\, \hat 1 \; \diamond \; \hat 2 \,u\, \hat 3 = \textit{undefined} \)

@@

\hspace{2mm}\(\sem{ \Diamond_a\,p } \hspace{4mm}{\triangleq} \hspace{4mm}\{x \mid \exists y . x{\hat{R}_a}y \wedge y \in \sem{ p }^{\epikspec}\} \)
  \caption{Interpretation of Epik terms as sets of guarded strings}
  \label{fig:interp2}
\end{figure}

@@

@@
